Fluent statistical computing interfaces for biological data analysis

### 2020-07-13 Rmd/04-intron.Rmd
partial spelling

Here we consider a tidy-data approach for exploring
estimated coverage from RNA-seq data. We establish a simple 
framework, for aggregating accross experimental design, and annotated genomic regions to discover 'interesting' coverage trace plots. We highlight how this framework can be used to
develop data descriptions that find putative genes with
intron retention. Our framework is implemented in a software
package called **superintronic**, available at https://github.com/sa-lee/superintronic.

@@

In high-throughput sequencing data sets, coverage is the estimated number of
reads that overlap a single position of the reference genome, and is important for assessing sequencing data quality and used in many different aspects of omic's analysis such peak-calling in ChIP-seq or varinat calling in DNA-seq [@Sims2014-qk]. Here we emphasise looking at coverage traces to find biological events of interest, rather than only relying on numerical summaries of the data. By facetting these traces over combinations of the experimental design and with their biological context such as gene annotations, we can gain an
insight into biological signal under study. 
Of course, due to the sheer size of most reference genome annotaions 
it would take an extremely long time for an analyst to look at every 
single possible region where there is some interesting signal within the
coverage trace. Because visualisation does not scale, we need to search the possible space of coverage traces and provide diagnostics for
identifying traces with interesting biology. We have taken inspiration from
visualisation literature; in particular the idea of scatterplot diagnostics ("scagnostics") for summarising the space of all possible 2D scatterplots
to a small number of descriptors of each scatterplots properties such as 
density or monotonicity [@Friedman2002-yx;@Wilkinson2005-xt].  Similarly,
searching for unusual time series via estimating descriptors such as seasonality or autocorrelation, and visualising those descriptors instead [@Hyndman2015-sc].

Although there are many flexible and powerful software tools, like BEDTools or DeepTools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantegous to have tooling
that is tightly coupled to a statistical computing language such as R. This allows interoperability between other software packages for data wrangling, visualisation and modelling within the ecosystem of the language that may not
be possible with a single command line tool. There are also gains in reproduciblity as analysts do not have to move between multiple software suites to explore their processed transcriptomics data.

@@

**superintronic** is an R package available on Github at https://github.com/sa-lee/superintronic and used for estimating, representing,
and visualising per-base coverage scores, that can then be 
flexibly summarised over factors within an experimental design and collapsed
over regions of the genome using our companion package **plyranges**. The
aspects of this combined workflow are summarised in figure \@ref(fig:si-fig).

@@

External reference annotations, perhaps transcripts or exons, 
can be coereced to **GRanges** objects, are  incorporated into the coverage **GRanges** by taking the intersection of
the annotation with the **GRanges** using an overlap intersect join from the **plyranges** software. The resulting intersection will now contain the
per base coverage that are overlapped the genomic features in the annotation,
along side any metadata about the features themselves. Since our main workflow
interest is in discovering coverage traces with IR profiles, **superintronic**
provides some syntatic sugar for unravelling gene annotations into their exonic
and intronic parts, and intersecting them with a coverage **GRanges**.

@@

Once the coverage **GRanges** has reference genomic features, data 
descriptors can be computed via collecting summary statistics across
the factors of the experimental design and features of interest. This can be
achieved using **plyranges** directly by first grouping across variables
of interest, computing descriptors defined by **superintronic** and then
pivoting the results into a wide form table for additional processing or
visualisation. There are many descriptors defined by **superintronic**
that are weighted statistics (as we have to account for the number of bases
covered, or the width of the range) of the coverage score, such as the mean 
and standard deviation. There are also descriptors that can be used to find
the number of times the coverage trace is above a certain number of bases
or score. To find coverage traces that have unusual descriptors, 
the descriptors can be visualised directly as a scatterplot matrix. After that
thresholds can be applied to filter the genomic features that had extreme
descriptors on the coverage **GRanges**, and the traces can be visualised.

@@

\begin{figure*}[!tpb]
\centerline{\includegraphics[width=\textwidth]{coverage-si-hits.pdf}}
\caption{{\it Superintronic} selects genes with IR-like coverage profiles in poly(A) RNA HCC827 cell lines. Genes (a) PSMB7, (b) EIF2S3, and (c) ARGLU1 are highlighted out of 43 genes selected. Whilst the analysis was carried out on poly(A) RNA libraries, coverage is shown for both poly(A) RNA HCC827 (top) and Total RNA HCC827 (bottom) samples to ensure that results are not an artifact of RNA library preparation. Coverage is oriented from 5' to 3', with exon regions coloured green and intron regions coloured orange.
}\label{fig:ir}
\end{figure*}

Using \textit{superintronic}, poly(A) RNA HCC827 cell lines were examined for genes with IR  after selecting genes in the hg38 reference that were protein coding, did not overlap any other gene and were placed on the main contigs -- a total of 6,606 genes. 
These genes were then split into intron and exon regions and intersected with the coverage of each sample. 
Per gene intron and exon summary statistics, mean and standard deviation, were computed on log-coverage values. 
We selected genes enriched for IR-like coverage profiles by looking for ``expressed'' genes, where for a substantial number of intron bases its coverage is much higher than other intron features within the same gene while having similar expression levels to the exon features. 
To do this we used the following thresholds -- genes had an average exon log-coverage of greater than 2 (corresponding to the mean of average exon log-coverage values across all genes), a standard deviation of intron log-coverage greater than 1.5 (corresponding to the mean of intron standard deviation values across all genes), and genes with a large number of intron bases with log-coverage greater than 2 (top 1\% of genes).
The thresholds were chosen after examining distributions of the summary statistics (Supplementary Figure 4).
Forty-three genes met these criteria, where a manual check of coverage profiles revealed that 36 genes indeed appear IR-like (see `Analyses' page in Supplementary Materials).
We highlight 3 of these genes in Figure~\ref{fig:ir}.
The coverage of remaining 7 genes appear to be more pre-mRNA-like, with large 
variation in intron coverage, where at its peak it is expressed similarly to exon features.

>>>>

Here we consider a tidy-data approach for exploring
estimated coverage from RNA-seq data. We establish a simple 
framework, for aggregating across experimental design, and annotated genomic regions to discover 'interesting' coverage trace plots. We highlight how this framework can be used to
develop data descriptions that find putative genes with
intron retention. Our framework is implemented in a software
package called **superintronic**, available at https://github.com/sa-lee/superintronic.

@@

In high-throughput sequencing data sets, coverage is the estimated number of
reads that overlap a single position of the reference genome, and is important for assessing sequencing data quality and used in many different aspects of omics analysis such peak-calling in ChIP-seq or variant calling in DNA-seq [@Sims2014-qk]. Here we emphasise looking at coverage traces to find biological events of interest, rather than only relying on numerical summaries of the data. By faceting these traces over combinations of the experimental design and with their biological context such as gene annotations, we can gain an
insight into biological signal under study. 
Of course, due to the sheer size of most reference genome annotations 
it would take an extremely long time for an analyst to look at every 
single possible region where there is some interesting signal within the
coverage trace. Because visualisation does not scale, we need to search the possible space of coverage traces and provide diagnostics for
identifying traces with interesting biology. We have taken inspiration from
visualisation literature; in particular the idea of scatter plot diagnostics ("scagnostics") for summarising the space of all possible 2D scatter plots
to a small number of descriptors of each scatter plots properties such as 
density or monotonicity [@Friedman2002-yx;@Wilkinson2005-xt].  Similarly,
searching for unusual time series via estimating descriptors such as seasonality or autocorrelation, and visualising those descriptors instead [@Hyndman2015-sc].

Although there are many flexible and powerful software tools, like BEDTools or DeepTools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantageous to have tooling
that is tightly coupled to a statistical computing language such as R. This allows interoperability between other software packages for data wrangling, visualisation and modelling within the ecosystem of the language that may not
be possible with a single command line tool. There are also gains in reproducibility as analysts do not have to move between multiple software suites to explore their processed transcriptomics data.

@@

**superintronic** is an R package available on GitHub at https://github.com/sa-lee/superintronic and used for estimating, representing,
and visualising per-base coverage scores, that can then be 
flexibly summarised over factors within an experimental design and collapsed
over regions of the genome using our companion package **plyranges**. The
aspects of this combined workflow are summarised in figure \@ref(fig:si-fig).

@@

External reference annotations, perhaps transcripts or exons, 
can be coerced to **GRanges** objects, are  incorporated into the coverage **GRanges** by taking the intersection of
the annotation with the **GRanges** using an overlap intersect join from the **plyranges** software. The resulting intersection will now contain the
per base coverage that are overlapped the genomic features in the annotation,
along side any metadata about the features themselves. Since our main workflow
interest is in discovering coverage traces with IR profiles, **superintronic**
provides some syntactic sugar for unravelling gene annotations into their exonic
and intronic parts, and intersecting them with a coverage **GRanges**.

@@

Once the coverage **GRanges** has reference genomic features, data 
descriptors can be computed via collecting summary statistics across
the factors of the experimental design and features of interest. This can be
achieved using **plyranges** directly by first grouping across variables
of interest, computing descriptors defined by **superintronic** and then
pivoting the results into a wide form table for additional processing or
visualisation. There are many descriptors defined by **superintronic**
that are weighted statistics (as we have to account for the number of bases
covered, or the width of the range) of the coverage score, such as the mean 
and standard deviation. There are also descriptors that can be used to find
the number of times the coverage trace is above a certain number of bases
or score. To find coverage traces that have unusual descriptors, 
the descriptors can be visualised directly as a scatter plot matrix. After that
thresholds can be applied to filter the genomic features that had extreme
descriptors on the coverage **GRanges**, and the traces can be visualised.

@@

\begin{figure*}[!tpb]
\centerline{\includegraphics[width=\textwidth]{coverage-si-hits.pdf}}
\caption{{\it Superintronic} selects genes with IR-like coverage profiles in poly(A) RNA HCC827 cell lines. Genes (a) PSMB7, (b) EIF2S3, and (c) ARGLU1 are highlighted out of 43 genes selected. Whilst the analysis was carried out on poly(A) RNA libraries, coverage is shown for both poly(A) RNA HCC827 (top) and Total RNA HCC827 (bottom) samples to ensure that results are not an artefact of RNA library preparation. Coverage is oriented from 5' to 3', with exon regions coloured green and intron regions coloured orange.
}\label{fig:ir}
\end{figure*}

Using \textit{superintronic}, poly(A) RNA HCC827 cell lines were examined for genes with IR  after selecting genes in the hg38 reference that were protein coding, did not overlap any other gene and were placed on the main configs -- a total of 6,606 genes. 
These genes were then split into intron and exon regions and intersected with the coverage of each sample. 
Per gene intron and exon summary statistics, mean and standard deviation, were computed on log-coverage values. 
We selected genes enriched for IR-like coverage profiles by looking for ``expressed'' genes, where for a substantial number of intron bases its coverage is much higher than other intron features within the same gene while having similar expression levels to the exon features. 
To do this we used the following thresholds -- genes had an average exon log-coverage of greater than 2 (corresponding to the mean of average exon log-coverage values across all genes), a standard deviation of intron log-coverage greater than 1.5 (corresponding to the mean of intron standard deviation values across all genes), and genes with a large number of intron bases with log-coverage greater than 2 (top 1\% of genes).
The thresholds were chosen after examining distributions of the summary statistics (Supplementary Figure 4).
Forty-three genes met these criteria, where a manual check of coverage profiles revealed that 36 genes indeed appear IR-like (see `Analyses' page in Supplementary Materials).
We highlight 3 of these genes in Figure~\ref{fig:ir}.
The coverage of remaining 7 genes appear to be more pre-mRNA-like, with large 
variation in intron coverage, where at its peak it is expressed similarly to exon features.
